---
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%", 
  cache.path = "cache/", 
  fig.height = 4, 
  fig.width = 7)
options(digits = 3)
```

# SimSurvNMarker (Simulate Survival process and Markers)
[![Build Status on Travis](https://travis-ci.org/boennecd/SimSurvNMarker.svg?branch=master,osx)](https://travis-ci.org/boennecd/SimSurvNMarker)

The `SimSurvNMarker` package reasonably fast simulates from a joint survival 
and marker model. Specifically, the package can simulate from the model 

<!-- $$\begin{align*} -->
<!-- \vec Y_{ij} \mid \vec U_i = \vec u_i -->
<!--   &\sim N^{(r)}(\vec \mu_i(s_{ij}, \vec u_i), \Sigma) -->
<!--   \\ -->
<!-- \vec\mu(s, \vec u) &= -->
<!--   \Gamma^\top \vec x_i + B^\top\vec g(s) + U^\top\vec m(s) -->
<!--   \\ -->
<!-- &= \left(I \otimes \vec x_i^\top\right)\text{vec}\Gamma -->
<!--      + \left(I \otimes \vec g(s)^\top\right)\text{vec} B -->
<!--      + \left(I \otimes \vec m(s)^\top\right) \vec u -->
<!--   \\ -->
<!-- \vec U_i &\sim N^{(K)}(\vec 0, \Psi) -->
<!--   \\ -->
<!-- h(t\mid \vec u) &= \exp\left( -->
<!--   \vec\omega^\top\vec b(t) + -->
<!--   \vec z_i^\top\vec\delta + -->
<!--   \vec\alpha^\top\vec\mu(t, \vec u) -->
<!--   \right) -->
<!--   \\ -->
<!-- &= \exp\Bigg( -->
<!--   \vec\omega^\top\vec b(t) + -->
<!--   \vec z_i^\top\vec\delta -->
<!--   + \vec 1^\top\left( -->
<!--   \text{diag}(\vec \alpha) \otimes \vec x_i^\top\right)\text{vec}\Gamma -->
<!--   + \vec 1^\top\left( -->
<!--   \text{diag}(\vec \alpha) \otimes \vec g(t)^\top\right)\text{vec} B \\ -->
<!-- &\hspace{50pt}+ \vec 1^\top\left( -->
<!--   \text{diag}(\vec \alpha) \otimes \vec m(t)^\top\right)\vec u -->
<!--   \Bigg) -->
<!-- \end{align*}$$ -->

$$\begin{align*}  \vec Y_{ij} \mid \vec U_i = \vec u_i    &\sim N^{(r)}(\vec \mu_i(s_{ij}, \vec u_i), \Sigma)    \\  \vec\mu(s, \vec u) &=    \Gamma^\top \vec x_i + B^\top\vec g(s) + U^\top\vec m(s)    \\  &= \left(I \otimes \vec x_i^\top\right)\text{vec}\Gamma       + \left(I \otimes \vec g(s)^\top\right)\text{vec} B       + \left(I \otimes \vec m(s)^\top\right) \vec u    \\  \vec U_i &\sim N^{(K)}(\vec 0, \Psi)    \\  h(t\mid \vec u) &= \exp\left(    \vec\omega^\top\vec b(t) +    \vec z_i^\top\vec\delta +    \vec\alpha^\top\vec\mu(t, \vec u)    \right)    \\  &= \exp\Bigg(    \vec\omega^\top\vec b(t) +    \vec z_i^\top\vec\delta    + \vec 1^\top\left(    \text{diag}(\vec \alpha) \otimes \vec x_i^\top\right)\text{vec}\Gamma    + \vec 1^\top\left(    \text{diag}(\vec \alpha) \otimes \vec g(t)^\top\right)\text{vec} B \\  &\hspace{50pt}+ \vec 1^\top\left(    \text{diag}(\vec \alpha) \otimes \vec m(t)^\top\right)\vec u    \Bigg)  \end{align*}$$

where $\vec Y_{ij}\in\mathbb R^{n_y}$ is individual $i$'s 
$j$th observed marker at time $s_{ij}$, 
$U_i\in\mathbb R^K$ is individual $i$'s random effect, and 
$h$ is the instantaneous hazard rate for the time-to-event outcome. 
$\vec\alpha$ is the so-called
association parameter. It shows the strength of the relation between the 
latent mean function, $\vec\mu(t,\vec u)$, and the log of the 
instantaneous rate, $h(t\mid \vec u)$. $\vec m(t)$, $\vec g(t)$ 
and $\vec b(t)$ are 
basis expansions of time. As an example, these can be a polynomial, 
a B-spline, or a natural cubic spline. The expansion for the baseline
hazard, $\vec b(t)$, is typically made on $\log t$ instead of $t$. 
One reason is that the model reduces to a Weibull distribution
when a first polynomial is used and $\vec\alpha = \vec 0$. $\vec x_i$ and 
$\vec z_i$ are individual specific known covariates.

We provide an example of how to use the package here and the 
[inst/test-data](inst/test-data) directory.

## Installation

The package is not on CRAN but can be installed through 

```{r how_to_install, eval = FALSE}
stopifnot(require(remotes)) # need the remotes package
install_github("boennecd/SimSurvNMarker")
```

## Example
We start by assigning all the variables that we will pass to the functions
in the package.

```{r attach_pkg}
library(SimSurvNMarker)
```

```{r model_params, cache = 1}
# quadrature nodes
gl_dat <- get_gl_rule(30L)

# spline functions
b_ks <- seq(log(1), log(10), length.out = 4)
m_ks <- seq(0, 10, length.out = 3)
g_ks <- m_ks

# simulation functions
r_n_marker <- function(id)
  rpois(1, 10) + 1L
r_obs_time <- function(id, n_markes)
  sort(runif(n_markes, 0, 10))
r_z <- function(id)
  as.numeric(runif(1L) > .5)
r_x <- function(id)
  numeric()
r_left_trunc <- function(id)
   rbeta(1, 1, 2) * 3
r_right_cens <- function(id)
  rbeta(1, 2, 1) * 6 + 4

# model parameters
omega <- c(-0.96, -2.26, -3.04, .45)
Psi <- structure(c(1.08, 0.12, -0.36, -0.48, 0.36, -0.12, 0.12, 0.36,
                   0, 0.12, 0, -0.12, -0.36, 0, 0.84, 0.12, 0.12, 0.12, -0.48, 0.12,
                   0.12, 0.84, -0.12, 0.24, 0.36, 0, 0.12, -0.12, 0.84, -0.12, -0.12,
                   -0.12, 0.12, 0.24, -0.12, 0.6), .Dim = c(6L, 6L))
B <- structure(c(0.97, 0.01, -0.07, -0.78, -0.86, 0.98), .Dim = 3:2)
sig <- diag(c(.2, .1)^2)
alpha <- c(0.7, 0.6)
delta <- 0.
gamma <- numeric()
```

```{r get_splines}
b_func <- get_ns_spline(b_ks, do_log = TRUE)
m_func <- get_ns_spline(m_ks, do_log = FALSE)
g_func <- get_ns_spline(g_ks, do_log = FALSE)
```

We show the baseline hazard function and the survival function without 
the marker below.

```{r plot_wo_marker}
# hazard function without marker
par(mar = c(5, 5, 1, 1))
plot(function(x) exp(drop(b_func(x) %*% omega)),
     xlim = c(1e-8, 10), ylim = c(0, .61), xlab = "Time",
     ylab = "Hazard (no marker)", xaxs = "i", bty = "l")

# survival function without marker
plot(function(x) eval_surv_base_fun(x, omega = omega, b_func = b_func, 
                                    gl_dat = gl_dat, delta = delta), 
     xlim = c(1e-4, 10),
     xlab = "Time", ylab = "Survival probability (no marker)", xaxs = "i",
     yaxs = "i", bty = "l", ylim = c(0, 1.01))
abline(h = .75, lty = 3)
abline(h = .25, lty = 3)
```

Next, we simulate individual specific markers. Each plot is for a given 
individual. The dashed curve is the mean, $\vec\mu(s, \vec 0)$, the 
fully drawn curve is the individual specific curve, $\vec\mu(s, \vec U_i)$,
the shaded areas are pointwise 95% interval for each mean curve, and the 
points are observed markers, $\vec y_{ij}$.

```{r show_sim_marker, fig.height=9, fig.width=9}
set.seed(1)
show_mark_mean <- function(B, Psi, sigma, m_func, g_func){
  tis <- seq(0, 10, length.out = 100)
  Psi_chol <- chol(Psi)
  
  par_old <- par(no.readonly = TRUE)
  on.exit(par(par_old))
  par(mar = c(4, 3, 1, 1), mfcol = c(4, 4))
  
  sigma_chol <- chol(sigma)
  n_y <- NCOL(sigma_chol)
  for(i in 1:16){
    U <- draw_U(Psi_chol, n_y = n_y)
    y_non_rng <- t(eval_marker(tis, B = B, g_func = g_func, U = NULL, 
                               offset = NULL, m_func = m_func))
    y_rng     <- t(eval_marker(tis, B = B, g_func = g_func, U = U, 
                               offset = NULL, m_func = m_func))

    sds <- sapply(tis, function(ti){
      M <- (diag(n_y) %x% m_func(ti))
      G <- (diag(n_y) %x% g_func(ti))
      sds <- sqrt(diag(tcrossprod(M %*% Psi, M)))
      cbind(drop(G %*% c(B)) - 1.96 * sds,
            drop(G %*% c(B)) + 1.96 * sds)
    }, simplify = "array")
    lbs <- t(sds[, 1, ])
    ubs <- t(sds[, 2, ])

    y_obs <- sim_marker(B = B, U = U, sigma_chol = sigma_chol, 
                        m_func = m_func, r_n_marker = r_n_marker, 
                        r_obs_time = r_obs_time, g_func = g_func, 
                        offset = NULL)

    matplot(tis, y_non_rng, type = "l", lty = 2, ylab = "", xlab = "Time",
            ylim = range(y_non_rng, y_rng, lbs, ubs, y_obs$y_obs))
    matplot(tis, y_rng    , type = "l", lty = 1, add = TRUE)
    matplot(y_obs$obs_time, y_obs$y_obs, type = "p", add = TRUE, pch = 3:4)

    polygon(c(tis, rev(tis)), c(lbs[, 1], rev(ubs[, 1])), border = NA,
            col = rgb(0, 0, 0, .1))
    polygon(c(tis, rev(tis)), c(lbs[, 2], rev(ubs[, 2])), border = NA,
            col = rgb(1, 0, 0, .1))

  }
  invisible()
}
show_mark_mean(B = B, Psi = Psi, sigma = sig, m_func = m_func, 
               g_func = g_func)
```

We sample a number of random effects and plot the hazard curves and survival 
functions given these random effects below. 

```{r show_draw_surv_curves, cache = 1, dependson = "model_params"}
set.seed(1)
local({
  par_old <- par(no.readonly = TRUE)
  on.exit(par(par_old))
  par(mfcol = c(1, 2))

  # hazard functions
  tis <- seq(1e-4, 10, length.out = 50)
  n_y <- NCOL(sig)
  Us <- replicate(100, draw_U(chol(Psi), n_y = n_y), 
                  simplify = "array")

  hz <- apply(Us, 3L, function(U)
    vapply(tis, function(x)
      exp(drop(delta + b_func(x) %*% omega +
                 alpha %*% eval_marker(ti = x, B = B, m_func = m_func, 
                                       g_func = g_func, U = U, 
                                       offset = NULL))),
      FUN.VALUE = numeric(1L)))

  matplot(tis, hz, lty = 1, type = "l", col = rgb(0, 0, 0, .1),
          xaxs = "i", bty = "l", yaxs = "i", ylim = c(0, 1),
          xlab = "time", ylab = "Hazard")

  # survival functions
  ys <- apply(Us, 3L, surv_func_joint,
              ti = tis, B = B, omega = omega, delta = delta,
              alpha = alpha, b_func = b_func, m_func = m_func, 
              gl_dat = gl_dat, g_func = g_func, offset = NULL)

  matplot(tis, ys, lty = 1, type = "l", col = rgb(0, 0, 0, .1),
          xaxs = "i", bty = "l", yaxs = "i", ylim = c(0, 1),
          xlab = "time", ylab = "Survival probability")
  abline(h = .75, lty = 3)
  abline(h = .25, lty = 3)
})
```

We end by drawing a data set. We consider the time it takes in seconds by 
using the `system.time` function. 

```{r sim_dat, cache = 1, dependson = "model_params"}
set.seed(70483614)
delta_vec <- 1
system.time(dat <- sim_joint_data_set(
  n_obs = 1000L, B = B, Psi = Psi, omega = omega, delta = delta_vec,
  alpha = alpha, sigma = sig, b_func = b_func, g_func = g_func,
  m_func = m_func, gl_dat = gl_dat, r_z = r_z, r_left_trunc = r_left_trunc, 
  r_right_cens = r_right_cens, r_n_marker = r_n_marker, 
  r_obs_time = r_obs_time, y_max = 10, gamma = gamma, r_x = r_x))
```

Finally, we show a few of the first rows along with some summary statistics.

```{r show_stats_sim_dat}
# survival data
head(dat$survival_data)

# marker data
head(dat$marker_data, 10)

# rate of observed events
mean(dat$survival_data$event) 

# mean event time
mean(subset(dat$survival_data, event           )$y)

# mean event time for the two group
mean(subset(dat$survival_data, event & Z1 == 1L)$y)
mean(subset(dat$survival_data, event & Z1 == 0L)$y)

# quantiles of the event time
quantile(subset(dat$survival_data, event)$y)

# fraction of observed markers per individual
NROW(dat$marker_data) / NROW(dat$survival_data)
```

## TODOs 

* Allow for derivatives of $\vec\mu$ in the hazard.  
* Provide example in the `man` files that needs it. 
