---
bibliography: ../ref.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%", 
  cache.path = "cache/", 
  fig.height = 4, 
  fig.width = 7, 
  dpi = 80)
options(digits = 3)
```

# SimSurvNMarker: Simulate Survival time and Markers
[![Build Status on Travis](https://travis-ci.org/boennecd/SimSurvNMarker.svg?branch=master,osx)](https://travis-ci.org/boennecd/SimSurvNMarker)
[![](https://www.r-pkg.org/badges/version/SimSurvNMarker)](https://CRAN.R-project.org/package=SimSurvNMarker)
[![CRAN RStudio mirror downloads](http://cranlogs.r-pkg.org/badges/SimSurvNMarker)](https://CRAN.R-project.org/package=SimSurvNMarker)

The `SimSurvNMarker` package reasonably fast simulates from a joint survival 
and marker model. The package uses a combination of Gauss–Legendre 
quadrature and one dimensional root finding to simulate the event times 
as described by @Crowther13. Specifically, the package can simulate from 
the model 

<!-- $$\begin{align*} -->
<!-- \vec Y_{ij} \mid \vec U_i = \vec u_i -->
<!--   &\sim N^{(r)}(\vec \mu_i(s_{ij}, \vec u_i), \Sigma) -->
<!--   \\ -->
<!-- \vec\mu_i(s, \vec u) &= -->
<!--   \Gamma^\top \vec x_i + B^\top\vec g(s) + U^\top\vec m(s) -->
<!--   \\ -->
<!-- &= \left(I \otimes \vec x_i^\top\right)\text{vec}\Gamma -->
<!--      + \left(I \otimes \vec g(s)^\top\right)\text{vec} B -->
<!--      + \left(I \otimes \vec m(s)^\top\right) \vec u -->
<!--   \\ -->
<!-- \vec U_i &\sim N^{(K)}(\vec 0, \Psi) -->
<!--   \\ -->
<!-- h(t\mid \vec u) &= \exp\left( -->
<!--   \vec\omega^\top\vec b(t) + -->
<!--   \vec z_i^\top\vec\delta + -->
<!--   \vec\alpha^\top\vec\mu_i(t, \vec u) -->
<!--   \right) -->
<!--   \\ -->
<!-- &= \exp\Bigg( -->
<!--   \vec\omega^\top\vec b(t) + -->
<!--   \vec z_i^\top\vec\delta -->
<!--   + \vec 1^\top\left( -->
<!--   \text{diag}(\vec \alpha) \otimes \vec x_i^\top\right)\text{vec}\Gamma -->
<!--   + \vec 1^\top\left( -->
<!--   \text{diag}(\vec \alpha) \otimes \vec g(t)^\top\right)\text{vec} B \\ -->
<!-- &\hspace{50pt}+ \vec 1^\top\left( -->
<!--   \text{diag}(\vec \alpha) \otimes \vec m(t)^\top\right)\vec u -->
<!--   \Bigg) -->
<!-- \end{align*}$$ -->

$$\begin{align*}\vec Y_{ij}\mid\vec U_i =\vec u_i &\sim N^{(r)}(\vec \mu_i(s_{ij}, \vec u_i), \Sigma)\\\vec\mu_i(s, \vec u)&=\Gamma^\top\vec x_i+B^\top\vec g(s)+U^\top\vec m(s)\\&=\left(I\otimes\vec x_i^\top\right)\text{vec}\Gamma+\left(I\otimes\vec g(s)^\top\right)\text{vec}B+\left(I\otimes\vec m(s)^\top\right)\vec u\\\vec U_i &\sim N^{(K)}(\vec 0,\Psi)\end{align*}$$

$$\begin{align*}h(t\mid\vec u)&=\exp\left(\vec\omega^\top\vec b(t)+\vec z_i^\top\vec\delta+\vec\alpha^\top\vec\mu_i(t, \vec u)\right)\\&=\exp\Bigg(\vec\omega^\top\vec b(t)+\vec z_i^\top\vec\delta+\vec 1^\top\left(\text{diag}(\vec\alpha)\otimes\vec x_i^\top\right)\text{vec}\Gamma+\vec 1^\top\left(\text{diag}(\vec\alpha)\otimes\vec g(t)^\top\right)\text{vec} B\\&\hspace{50pt}+\vec 1^\top\left(\text{diag}(\vec\alpha)\otimes\vec m(t)^\top\right)\vec u\Bigg)\end{align*}$$

where $\vec Y_{ij}\in\mathbb R^{n_y}$ is individual $i$'s 
$j$th observed marker at time $s_{ij}$, 
$U_i\in\mathbb R^K$ is individual $i$'s random effect, and 
$h$ is the instantaneous hazard rate for the time-to-event outcome. 
$\vec\alpha$ is the so-called
association parameter. It shows the strength of the relation between the 
latent mean function, $\vec\mu_i(t,\vec u)$, and the log of the 
instantaneous rate, $h(t\mid \vec u)$. $\vec m(t)$, $\vec g(t)$ 
and $\vec b(t)$ are 
basis expansions of time. As an example, these can be a polynomial, 
a B-spline, or a natural cubic spline. The expansion for the baseline
hazard, $\vec b(t)$, is typically made on $\log t$ instead of $t$. 
One reason is that the model reduces to a Weibull distribution
when a first polynomial is used and $\vec\alpha = \vec 0$. $\vec x_i$ and 
$\vec z_i$ are individual specific known time-invariant covariates.

We provide an example of how to use the package here, in 
[inst/test-data](https://github.com/boennecd/SimSurvNMarker/tree/master/inst/test-data) 
directory on Github, and at 
[rpubs.com/boennecd/SimSurvNMarker-ex](https://rpubs.com/boennecd/SimSurvNMarker-ex)
where we show that we simulate from the correct model by using a 
simulation study. The former examples includes

 - using [polynomials](#example-polynomial) as the basis functions.
 - using [polynomials with derivatives in the log hazard](#using-derivatives) as the basis functions.
 - using [natural cubic splines](#example-natural-cubic-splines) as the basis functions.

The purpose/goal of this package is to

 - allow the user to specify all components in R including the basis 
   functions. 
 - simulate reasonably fast from the model given this requirement.

## Installation

The package can be installed from Github using the`remotes` package:

```{r how_to_install, eval = FALSE}
stopifnot(require(remotes)) # needs the remotes package
install_github("boennecd/SimSurvNMarker")
```

It can also be installed from CRAN using `install.packages`: 

```{r CRAN_how_to_install, eval = FALSE}
install.packages("SimSurvNMarker")
```

## Example with Polynomials
We start with an example where we use polynomials as the basis functions. 
First, we assign the polynomial functions we will use. 

```{r poly_assign}
b_func <- function(x){
  x <- x - 1
  cbind(x^3, x^2, x)
}
g_func <- function(x){
  x <- x - 1
  cbind(x^3, x^2, x)
}
m_func <- function(x){
  x <- x - 1
  cbind(x^2, x, 1)
}
```

We use a third order polynomial for the two fixed terms, $\vec b$ and 
$\vec g$, and a second order random polynomial for the random term, 
$U_i^\top \vec m(s)$, in the latent mean function. We choose the 
following parameters for the baseline hazard.

```{r attach_pkg}
library(SimSurvNMarker)
```

```{r poly_base_haz}
omega <- c(1.4, -1.2, -2.1)
delta <- -.5 # the intercept

# quadrature nodes
gl_dat <- get_gl_rule(30L)

# hazard function without marker
par(mar = c(5, 5, 1, 1), mfcol = c(1, 2))
plot(function(x) exp(drop(b_func(x) %*% omega) + delta),
     xlim = c(0, 2), ylim = c(0, 1.5), xlab = "Time",
     ylab = "Hazard (no marker)", xaxs = "i",  yaxs = "i", bty = "l")
grid()

# survival function without marker
plot(function(x) eval_surv_base_fun(x, omega = omega, b_func = b_func, 
                                    gl_dat = gl_dat, delta = delta), 
     xlim = c(1e-4, 2.5),
     xlab = "Time", ylab = "Survival probability (no marker)", xaxs = "i",
     yaxs = "i", bty = "l", ylim = c(0, 1.01))
grid()
```

Then we set the following parameters for the random effect, $\vec U_i$, and
the parameters for the marker process. We also simulate a number of latent
markers' mean curves and observed marker values and plot the result. 
The dashed curve is the mean, $\vec\mu_i(s, \vec 0)$, the 
fully drawn curve is the individual specific curve, 
$\vec\mu_i(s, \vec U_i)$,
the shaded areas are a pointwise 95% interval for each mean curve, and the 
points are observed markers, $\vec y_{ij}$.

```{r poly_marker_setup, fig.height=9, fig.width=9}
r_n_marker <- function(id)
  # the number of markers is Poisson distributed
  rpois(1, 10) + 1L
r_obs_time <- function(id, n_markes)
  # the observations times are uniform distributed 
  sort(runif(n_markes, 0, 2))

Psi <- structure(c(0.18, 0.05, -0.05, 0.1, -0.02, 0.06, 0.05, 0.34, -0.25, 
                   -0.06, -0.03, 0.29, -0.05, -0.25, 0.24, 0.04, 0.04, 
                   -0.12, 0.1, -0.06, 0.04, 0.34, 0, -0.04, -0.02, -0.03, 
                   0.04, 0, 0.1, -0.08, 0.06, 0.29, -0.12, -0.04, -0.08, 
                   0.51), .Dim = c(6L, 6L))
B <- structure(c(-0.57, 0.17, -0.48, 0.58, 1, 0.86), .Dim = 3:2)
sig <- diag(c(.6, .3)^2)

# function to simulate a given number of individuals' markers' latent means
# and observed values
show_mark_mean <- function(B, Psi, sigma, m_func, g_func, ymax){
  tis <- seq(0, ymax, length.out = 100)
  Psi_chol <- chol(Psi)
  
  par_old <- par(no.readonly = TRUE)
  on.exit(par(par_old))
  par(mar = c(4, 3, 1, 1), mfcol = c(4, 4))
  
  sigma_chol <- chol(sigma)
  n_y <- NCOL(sigma_chol)
  for(i in 1:16){
    U <- draw_U(Psi_chol, n_y = n_y)
    y_non_rng <- t(eval_marker(tis, B = B, g_func = g_func, U = NULL, 
                               offset = NULL, m_func = m_func))
    y_rng     <- t(eval_marker(tis, B = B, g_func = g_func, U = U, 
                               offset = NULL, m_func = m_func))

    sds <- sapply(tis, function(ti){
      M <- (diag(n_y) %x% m_func(ti))
      G <- (diag(n_y) %x% g_func(ti))
      sds <- sqrt(diag(tcrossprod(M %*% Psi, M)))
      cbind(drop(G %*% c(B)) - 1.96 * sds,
            drop(G %*% c(B)) + 1.96 * sds)
    }, simplify = "array")
    lbs <- t(sds[, 1, ])
    ubs <- t(sds[, 2, ])

    y_obs <- sim_marker(B = B, U = U, sigma_chol = sigma_chol, 
                        m_func = m_func, r_n_marker = r_n_marker, 
                        r_obs_time = r_obs_time, g_func = g_func, 
                        offset = NULL)

    matplot(tis, y_non_rng, type = "l", lty = 2, ylab = "", xlab = "Time",
            ylim = range(y_non_rng, y_rng, lbs, ubs, y_obs$y_obs))
    matplot(tis, y_rng    , type = "l", lty = 1, add = TRUE)
    matplot(y_obs$obs_time, y_obs$y_obs, type = "p", add = TRUE, pch = 3:4)

    polygon(c(tis, rev(tis)), c(lbs[, 1], rev(ubs[, 1])), border = NA,
            col = rgb(0, 0, 0, .1))
    polygon(c(tis, rev(tis)), c(lbs[, 2], rev(ubs[, 2])), border = NA,
            col = rgb(1, 0, 0, .1))

  }
  invisible()
}
set.seed(1)
show_mark_mean(B = B, Psi = Psi, sigma = sig, m_func = m_func, 
               g_func = g_func, ymax = 2)
```

As an example, we simulate the random effects and plot the conditional 
hazards and survival functions. We start by assigning the association 
parameter, $\vec\alpha$.

```{r poly_show_cond_surv}
alpha <- c(.5, .9)

# function to plot simulated conditional hazards and survival 
# functions
sim_surv_curves <- function(sig, Psi, delta, omega, alpha, B, m_func, 
                            g_func, b_func, ymax) {
  par_old <- par(no.readonly = TRUE)
  on.exit(par(par_old))
  par(mfcol = c(1, 2), mar = c(5, 5, 1, 1))

  # hazard functions
  tis <- seq(0, ymax, length.out = 50)
  n_y <- NCOL(sig)
  Us <- replicate(100, draw_U(chol(Psi), n_y = n_y), 
                  simplify = "array")

  hz <- apply(Us, 3L, function(U)
    vapply(tis, function(x)
      exp(drop(delta + b_func(x) %*% omega +
                 alpha %*% eval_marker(ti = x, B = B, m_func = m_func, 
                                       g_func = g_func, U = U, 
                                       offset = NULL))),
      FUN.VALUE = numeric(1L)))

  matplot(tis, hz, lty = 1, type = "l", 
          col = rgb(0, 0, 0, .1), xaxs = "i", bty = "l", yaxs = "i", 
          ylim = c(0, max(hz, na.rm = TRUE)), xlab = "time", 
          ylab = "Hazard")
  grid()

  # survival functions
  ys <- apply(Us, 3L, surv_func_joint,
              ti = tis, B = B, omega = omega, delta = delta,
              alpha = alpha, b_func = b_func, m_func = m_func, 
              gl_dat = gl_dat, g_func = g_func, offset = NULL)

  matplot(tis, ys, lty = 1, type = "l", col = rgb(0, 0, 0, .1),
          xaxs = "i", bty = "l", yaxs = "i", ylim = c(0, 1),
          xlab = "time", ylab = "Survival probability")
  grid()
}

set.seed(1)
sim_surv_curves(sig = sig, Psi = Psi, delta = delta, omega = omega, 
                alpha = alpha, B = B, m_func = m_func, g_func = g_func, 
                b_func = b_func, ymax = 2)
```

We end by assigning the functions to get the covariates, coefficients for 
the fixed effects, the left-truncation function, and right-censoring 
function. 

```{r poly_assign_residuals}
r_z <- function(id)
  # returns a design matrix for a dummy setup
  cbind(1, (id %% 3) == 1, (id %% 3) == 2)
r_z(1:6) # covariates for the first six individuals

# same covariates for the fixed time-invariant effects for the marker
r_x <- r_z

r_left_trunc <- function(id)
  # no left-truncation
  0
r_right_cens <- function(id)
  # right-censoring time is exponentially distributed
  rexp(1, rate = .5)

# fixed effect coefficients in the hazard
delta_vec <- c(delta, -.5, .5)
# fixed time-invariant effect coefficients in the marker process 
gamma <- matrix(c(.25, .5, 0, -.4, 0, .3), 3, 2)
```

A full data set can now be simulated as follows. We consider the time it 
takes in seconds by using the `system.time` function. 

```{r poly_sim_full, cache = 1}
set.seed(70483614)
system.time(dat <- sim_joint_data_set(
  n_obs = 1000L, B = B, Psi = Psi, omega = omega, delta = delta_vec,
  alpha = alpha, sigma = sig, b_func = b_func, g_func = g_func,
  m_func = m_func, gl_dat = gl_dat, r_z = r_z, r_left_trunc = r_left_trunc, 
  r_right_cens = r_right_cens, r_n_marker = r_n_marker, 
  r_obs_time = r_obs_time, y_max = 2, gamma = gamma, r_x = r_x))
```

The first entries of the survival data and the observed markers looks as 
follows.

```{r poly_show_stats_sim_dat}
# survival data
head(dat$survival_data)

# marker data
head(dat$marker_data, 10)
```

To illustrate that we simulate from the correct model, we can estimate a 
linear mixed models for the markers as follows. 

```{r poly_est_lme, cache = 1, dependson = "poly_sim_full"}
library(lme4)

# estimate the linear mixed model (skip this if you want and look at the 
# estimates in the end)
local({
  m_dat <- dat$marker_data
  n_y <- NCOL(gamma)
  d_x <- NROW(gamma)
  d_g <- NROW(B)
  d_m <- NROW(Psi) / n_y
  
  Y_names <- paste0("Y", 1:n_y)
  id_vars <- c("id", "obs_time")
  if(d_x > 0)
    id_vars <- c(id_vars, paste0("X", seq_len(d_x)))
  
  lme_dat <- melt(m_dat, id.vars = id_vars, measure.vars = Y_names, 
                  variable.name = "XXTHEVARIABLEXX", 
                  value.name = "XXTHEVALUEXX")
  
  if(length(alpha) > 1){
    if(length(B) > 0L)
      frm <- substitute(
        XXTHEVALUEXX ~
          XXTHEVARIABLEXX : g_func(ti) - 1L +
          (XXTHEVARIABLEXX : m_func(ti) - 1L | i),
        list(ti = as.name("obs_time"), i = as.name("id")))
    else 
      frm <- substitute(
        XXTHEVALUEXX ~
          (XXTHEVARIABLEXX : m_func(ti, m_ks) - 1L | i),
        list(ti = as.name("obs_time"), i = as.name("id")))
    frm <- eval(frm)
    
    if(d_x > 0)
      for(i in rev(seq_len(d_x))){
        frm_call <- substitute(
          update(frm, . ~ XXTHEVARIABLEXX : x_var + .),
          list(x_var = as.name(paste0("X", i))))
        frm <- eval(frm_call)
      }
    
  } else {
    if(length(B) > 0L)
      frm <- substitute(
        XXTHEVALUEXX ~
          ns_func(ti, g_ks) - 1L +
          (ns_func(ti, m_ks) - 1L | i),
        list(ti = as.name("obs_time"), i = as.name("id"), 
             g_ks = as.name("g_ks"), m_ks = as.name("m_ks")))
    else 
      frm <- substitute(
        XXTHEVALUEXX ~
          (ns_func(ti, m_ks) - 1L | i),
        list(ti = as.name("obs_time"), i = as.name("id"), 
             m_ks = as.name("m_ks")))
    frm <- eval(frm)
    
    if(d_x > 0)
      for(i in rev(seq_len(d_x))){
        frm_call <- substitute(
          update(frm, . ~ x_var + .),
          list(x_var = as.name(paste0("X", i))))
        frm <- eval(frm_call)
      }
  }
        
  fit <- lmer(frm, lme_dat, control = lmerControl(
    check.conv.grad = .makeCC("ignore", tol = 1e-3, relTol = NULL)))
  
  gamma <- t(matrix(fixef(fit)[seq_len(d_x * n_y)], nr = n_y))
  
  B <- t(matrix(fixef(fit)[seq_len(d_g * n_y) + (d_x * n_y)], nr = n_y))
  vc <- VarCorr(fit)
  Psi <- vc$id
  attr(Psi, "correlation") <- attr(Psi, "stddev") <- NULL
  dimnames(Psi) <- NULL
  K <- SimSurvNMarker:::get_commutation(n_y, d_m)
  Psi <- tcrossprod(K %*% Psi, K)

  Sigma <- diag(attr(vc, "sc")^2, n_y)

  list(gamma = gamma, B = B, Psi = Psi, Sigma = Sigma)
})
```

Although we assume equal noise variance, the estimates are close to the 
true values.

```{r comp_poly_est_lme}
gamma
B
Psi
sig
```

We then fit a Cox model with only the observed markers (likely biased but gives us an 
idea about whether we are using the correct model).

```{r poly_est_coxph}
local({
  library(survival)
  tdat <- tmerge(dat$survival_data, dat$survival_data, id = id, 
                 tstart = left_trunc, tstop = y, ev = event(y, event))
  
  for(i in seq_along(alpha)){
    new_call <- substitute(tmerge(
      tdat, dat$marker_data, id = id, tdc(obs_time, YVAR)),
      list(YVAR = as.name(paste0("Y", i))))
    names(new_call)[length(new_call)] <- paste0("Y", i)
    tdat <- eval(new_call)
  }
  tdat <- na.omit(tdat)
  
  sformula <- Surv(left_trunc, y, ev) ~ 1
  for(i in seq_along(delta_vec)){
    new_call <- substitute(update(sformula, . ~ . + XVAR), 
                           list(XVAR = as.name(paste0("Z", i))))
    sformula <- eval(new_call)
  }
  for(i in seq_along(alpha)){
    new_call <- substitute(update(sformula, . ~ . + XVAR), 
                           list(XVAR = as.name(paste0("Y", i))))
    sformula <- eval(new_call)
  }
  
  fit <- coxph(sformula, tdat)
  print(summary(fit))  
  invisible(fit)
})
```

This is close-ish to the true values.

```{r comp_poly_est_coxph}
delta_vec
alpha
```

### Using Derivatives
It is possible to use derivatives of the latent mean, 
$\vec\mu_i(s, \vec u)$, with respect to time in the hazard. As an example, 
we consider the first-order derivative below and plot conditional hazards
and survival functions simulated from the new model. 

```{r poly_derivs}
g_func_surv <- function(x){
  x <- x - 1
  cbind(3 * x^2, 2 * x, 1)
}
m_func_surv <- function(x){
  x <- x - 1
  cbind(2 * x, 1, 0)
}

set.seed(1)
sim_surv_curves(sig = sig, Psi = Psi, delta = delta, omega = omega, 
                alpha = alpha, B = B, m_func = m_func_surv, 
                g_func = g_func_surv, b_func = b_func, ymax = 2)
```

A new data set can now be simulated as follows. 

```{r get_data_poly_derivs, cache = 1}
set.seed(70483614)
system.time(dat <- sim_joint_data_set(
  n_obs = 1000L, B = B, Psi = Psi, omega = omega, delta = delta_vec,
  alpha = alpha, sigma = sig, b_func = b_func, g_func = g_func,
  m_func = m_func, gl_dat = gl_dat, r_z = r_z, r_left_trunc = r_left_trunc, 
  r_right_cens = r_right_cens, r_n_marker = r_n_marker, 
  r_obs_time = r_obs_time, y_max = 2, gamma = gamma, r_x = r_x, 
  # the additions
  m_func_surv = m_func_surv, g_func_surv = g_func_surv, 
  use_fixed_latent = FALSE))
```

The first entries of the new data looks as follows.

```{r derivs_poly_show_stats_sim_dat}
# survival data
head(dat$survival_data)

# marker data
head(dat$marker_data, 10)
```

```{r poly_clean_up, echo = FALSE}
rm(list = ls()[!ls() %in% c("sim_surv_curves", "show_mark_mean")])
```

## Example with Natural Cubic Splines
In this section, we will use natural cubic splines for the time-varying 
basis functions. We start by assigning all the variables that we will pass 
to the functions in the package.

```{r model_params, cache = 1}
# quadrature nodes
gl_dat <- get_gl_rule(30L)

# knots for the spline functions
b_ks <- seq(log(1), log(10), length.out = 4)
m_ks <- seq(0, 10, length.out = 3)
g_ks <- m_ks

# simulation functions
r_n_marker <- function(id)
  rpois(1, 10) + 1L
r_obs_time <- function(id, n_markes)
  sort(runif(n_markes, 0, 10))
r_z <- function(id)
  as.numeric(runif(1L) > .5)
r_x <- function(id)
  numeric()
r_left_trunc <- function(id)
   rbeta(1, 1, 2) * 3
r_right_cens <- function(id)
  rbeta(1, 2, 1) * 6 + 4

# model parameters
omega <- c(-0.96, -2.26, -3.04, .45)
Psi <- structure(c(1.08, 0.12, -0.36, -0.48, 0.36, -0.12, 0.12, 0.36,
                   0, 0.12, 0, -0.12, -0.36, 0, 0.84, 0.12, 0.12, 0.12, -0.48, 0.12,
                   0.12, 0.84, -0.12, 0.24, 0.36, 0, 0.12, -0.12, 0.84, -0.12, -0.12,
                   -0.12, 0.12, 0.24, -0.12, 0.6), .Dim = c(6L, 6L))
B <- structure(c(0.97, 0.01, -0.07, -0.78, -0.86, 0.98), .Dim = 3:2)
sig <- diag(c(.2, .1)^2)
alpha <- c(0.7, 0.6)
delta <- 0.
gamma <- numeric()
```

```{r get_splines}
# spline functions
b_func <- get_ns_spline(b_ks, do_log = TRUE)
m_func <- get_ns_spline(m_ks, do_log = FALSE)
g_func <- get_ns_spline(g_ks, do_log = FALSE)
```

Notice that we use the `get_ns_spline` functions which reduces the 
computation time a lot. We show the baseline hazard function and the 
survival function without the marker below.

```{r plot_wo_marker}
# hazard function without marker
par(mar = c(5, 5, 1, 1), mfcol = c(1, 2))
plot(function(x) exp(drop(b_func(x) %*% omega)),
     xlim = c(1e-8, 10), ylim = c(0, .61), xlab = "Time",
     ylab = "Hazard (no marker)", xaxs = "i", bty = "l")
grid()

# survival function without marker
plot(function(x) eval_surv_base_fun(x, omega = omega, b_func = b_func, 
                                    gl_dat = gl_dat, delta = delta), 
     xlim = c(1e-4, 10),
     xlab = "Time", ylab = "Survival probability (no marker)", xaxs = "i",
     yaxs = "i", bty = "l", ylim = c(0, 1.01))
grid()
```

Next, we simulate individual specific markers. Each plot is for a given 
individual.

```{r show_sim_marker, fig.height=9, fig.width=9}
set.seed(1)
show_mark_mean(B = B, Psi = Psi, sigma = sig, m_func = m_func, 
               g_func = g_func, ymax = 10)
```

We sample a number of random effects and plot the hazard curves and survival 
functions given these random effects below. 

```{r show_draw_surv_curves}
set.seed(1)
sim_surv_curves(sig = sig, Psi = Psi, delta = delta, omega = omega, 
                alpha = alpha, B = B, m_func = m_func, g_func = g_func, 
                b_func = b_func, ymax = 10)
```

We end by drawing a data set.

```{r sim_dat, cache = 1, dependson = "model_params"}
set.seed(70483614)
delta_vec <- 1
system.time(dat <- sim_joint_data_set(
  n_obs = 1000L, B = B, Psi = Psi, omega = omega, delta = delta_vec,
  alpha = alpha, sigma = sig, b_func = b_func, g_func = g_func,
  m_func = m_func, gl_dat = gl_dat, r_z = r_z, r_left_trunc = r_left_trunc, 
  r_right_cens = r_right_cens, r_n_marker = r_n_marker, 
  r_obs_time = r_obs_time, y_max = 10, gamma = gamma, r_x = r_x))
```

Finally, we show a few of the first rows along with some summary statistics.

```{r show_stats_sim_dat}
# survival data
head(dat$survival_data)

# marker data
head(dat$marker_data, 10)

# rate of observed events
mean(dat$survival_data$event) 

# mean event time
mean(subset(dat$survival_data, event           )$y)

# mean event time for the two group
mean(subset(dat$survival_data, event & Z1 == 1L)$y)
mean(subset(dat$survival_data, event & Z1 == 0L)$y)

# quantiles of the event time
quantile(subset(dat$survival_data, event)$y)

# fraction of observed markers per individual
NROW(dat$marker_data) / NROW(dat$survival_data)
```

## References
